---
title: ""
subtitle: ""
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    keep_tex: true
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

## Stat 242 Quiz -- Topics Drawn from Sections 5.5 and Chapter 3

## What's Your Name? ____________________

In each of the following data problems there is a potential violation of the assumption of independence.  Explain what the potential problem is in a sentence or two.

#### 1. Researchers interested in learning the effects of speed limits on traffic accidents recorded the number of accidents per year for each of 10 consecutive years on roads in a state with speed limits of 90 km/h.  They also recorded the number of accidents for the next 7 years on the same roads after the speed limit had been increased to 110 km/h.  They conducted statistical analyses to compare the mean number of accidents per year when the speed limit was 90 km/h and the mean number of accidents per year when the speed limit was 110 km/h.

\vspace{6cm}

#### 2. Researchers interested in investigating the effect of indoor pollution on respiratory health randomly selected houses in a particular city.  Each house was monitored for nitrogen dioxide concentration and categorized as being either high or low on the nitrogen dioxide scale.  Each member of the household was measured for respiratory health in terms of breathing capacity.  They conducted statistical analyses to compare the mean breathing capacity score for people in houses with high nitrogen dioxide levels and the mean breathing capacity score for people in houses with low nitrogen dioxide levels.

\newpage

(this page left intentionally blank)

\newpage

## Stat 242 Quiz -- Topics Drawn from Sections 5.5 and Chapter 3

## What's Your Name? ____________________

We have data for the relative brain weights (brain weight divided by body weight) for 51 species of mammal whose average litter size is less than 2, and for 45 species of mammal whose average litter size is greater than or equal to 2.  (Across both groups, we have data for 96 species.)

The following R code displays the first few rows of the data set, plots of the data, and the standard deviations of the relative brain weights for the observations within each group.

```{r, message = FALSE, echo = FALSE, warning=FALSE}
library(Sleuth3)
library(gridExtra)
library(ggplot2)
library(dplyr)

brain_sizes <- ex0333
```

```{r, echo = TRUE, fig.height = 4}
head(brain_sizes)
```

```{r, fig.height = 3}
ggplot(data = brain_sizes, mapping = aes(x = BrainSize, color = LitterSize)) +
  geom_density()


brain_sizes %>%
  group_by(LitterSize) %>%
  summarize(
    sd_brain_size = sd(BrainSize)
  )
```

\newpage

#### 1. Discuss any conditions for the anova model that are not satisfied.

\vspace{9cm}

#### 2. Suggest a strategy that could be used to help address the issues you identified in problem 1.
