---
title: "R Code: F tests for ANOVA, More Nested Models"
subtitle: "(Sleuth3 Sections 5.3 and 5.4)"
date: "2019-09-22"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    highlight: zenburn
header-includes:
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE)
```

## Spock Trials Example

```{r, message = FALSE, echo = FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
library(gmodels)

juries <- read_csv("http://www.evanlray.com/data/sleuth3/ex0502_women_jurors.csv")

ggplot(data = juries, mapping = aes(x = Judge, y = Percent)) +
  geom_boxplot()
```

### Question: are the means for judges A through F equal?

$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5 = \mu_6$

$H_A:$ At least one mean for a judge other than Spock is not equal to the others.

Note: if the null hypothesis is correct, there are effectively two groups:

1. All judges other than Spock's (they all have the same mean!)
2. Spock's judge

### Tasks we need to do:

1. Fit the full model
2. Create a new variable in the data set representing the groups in the reduced model
3. Fit the reduced model
4. Call `anova` to compare the reduced model to the full model

\newpage

#### Step 1: Fit the full model

```{r}
fit_full <- lm(Percent ~ Judge, data = juries)
summary(fit_full)
```

#### Step 2: Create a new variable in the data set representing the groups in the reduced model

The mutate statement below creates a new variable in the `juries` data frame called `judge_reduced`.  For each observation in the data set, `judge_reduced` is `"Other"` if the Judge was in `"A"`, `"B"`, `"C"`, `"D"`, `"E"`, or `"F"`.  Otherwise, `judge_reduced` is `"Spock's"`.

```{r}
juries <- juries %>%
  mutate(
    judge_reduced = ifelse(Judge %in% c("A", "B", "C", "D", "E", "F"), "Other", "Spock's")
  )

head(juries, 15) # just to check and make sure our new variable was created correctly
```

#### Step 3: Fit the reduced model

```{r}
fit_reduced <- lm(Percent ~ judge_reduced, data = juries)
summary(fit_reduced)
```

#### Step 4: Call `anova` to compare the reduced model to the full model

```{r}
anova(fit_reduced, fit_full)
```


#### Not part of the R code, but I can't resist: what is the conclusion?  Does this prove that the other judges had the same mean percent of women in their jury pools?


\blandscape
\includegraphics{F_examples_p1.pdf}

\includegraphics{F_examples_p2.pdf}
\elandscape

