---
title: "\"Simple\" Linear Regression"
subtitle: "20191004 -- Sleuth 3 Chapter 7"
output: pdf_document
header-includes:
   - \usepackage{soul}
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
```

## Example

We have a data set with information about 152 flights by Endeavour Airlines that departed from JFK airport in New York to either Nashville (BNA), Cincinnati (CVG), or Minneapolis-Saint Paul (MSP) in January 2012.

```{r, echo=FALSE, message = FALSE}
flights <- read_csv("flights_data.csv")
```

```{r}
head(flights, 4)
```

## So Far: ANOVA Model

 * Observations in group $i$ follow a Normal($\mu_i$, $\sigma^2$) distribution
 * Observations are independent of each other

```{r, fig.height = 1.5}
ggplot(data = flights, mapping = aes(x = air_time, color = dest)) +
  geom_density() +
  theme_bw()
```

**Note:** The picture would look exactly the same if we treated distance as a categorical variable:

```{r, fig.height = 1.5}
ggplot(data = flights, mapping = aes(x = air_time, color = factor(distance))) +
  geom_density() +
  theme_bw()
```

\newpage

**Old idea:** Each group has a normal distribution with its own mean

 * Categorical explanatory variable

**New idea:** Each group has a normal distribution with a mean that is a linear function of distance

 * Quantitative (numeric) explanatory variable

```{r, echo = FALSE, message=FALSE, fig.height = 3}
library(ggridges)

lm_fit <- lm(air_time ~ distance, data = flights)

ggplot(data = flights, mapping = aes(x = air_time, y = distance)) +
  geom_abline(intercept = -(coef(lm_fit)[1])/(coef(lm_fit)[2]), slope = 1/coef(lm_fit)[2]) +
  geom_density_ridges(mapping = aes(fill = factor(distance)), alpha = 0.4, jittered_points = TRUE, position = "points_jitter") +
  scale_fill_discrete("distance") +
  coord_flip() +
  theme_ridges(center = TRUE)
```

The simple linear regression is exactly like the ANOVA model, with the one new restriction that the means fall along a line.

### Two ways to write the model:

**Focusing on the mean (book)**

Values of the response variable are independent and normally distributed with mean $\mu(Y|X) = \beta_0 + \beta_1 X$

  * Read as "The mean of Y for a given value of X"
  * In our example, Y is air time and X is distance.

**Written for a single observation, number $i$ (my preference)**

\vspace{-0.5cm}

\begin{align*}
Y_i &= \beta_0 + \beta_1 x_i + \varepsilon_i \\
\varepsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

 * In our example, $Y_i$ is the air time for flight number $i$ and $x_i$ is the distance for flight number $i$.

**Parameter interpretations**

* $\beta_0$ is intercept for the population: mean value of the response when $X = 0$, in the population
* $\beta_1$ is slope for the population: change in mean response when $X$ increases by 1 unit, in the population.
* $\beta_0$ and $\beta_1$ are unknown population paramters.  We estimate them with the intercept and slope of a line describing our sample.

### Conditions: spells "LINE-O"

Exactly the same as conditions for ANOVA, with addition that the mean of the response is a linear function of the explanatory varible:

 * **Linear** relationship between explanatory and response variables
 * **Independent** observations (knowing that one observation is above its mean wouldn't give you any information about whether or not another observation is above its mean)
 * **Normal** distribution
 * **Equal standard deviation** of response for all values of X
    * Denote this standard deviation by $\sigma$
 * **no Outliers** (not a formal part of the model, but important to check in practice)
