---
title: "p-values for t tests using the pt function in R"
output: pdf_document
geometry: margin=0.6in
---

## Two things `pt` can calculate:

Suppose our calculated t statistic is -2, and the degrees of freedom is 5.

#### We can calculate the probability of obtaining a t statistic less than or equal to -2 as follows:

```{r}
pt(-2, df = 5)
```

This is the area under the $t$ distribution density curve to the left of -2.

```{r, echo = FALSE, message=FALSE, fig.height=2}
library(ggplot2)
library(grid)

x_grid_1 <- seq(from = -4, to = -2, length = 101)

region_to_shade1 <- data.frame(
  x = c(-4, x_grid_1, -2),
  y = c(0, dt(x_grid_1, df = 14, log = FALSE), 0)
)

ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) +
  stat_function(fun = dt, args = list(df = 14)) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = -2) +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.4), expand = FALSE) +
  theme_bw() +
  xlab("t") +
  ylab("t distribution density") +
  ggtitle(expression(paste("Shaded area is ", P(t <= -2))))
```

#### We can calculate the probability of obtaining a t statistic greater than or equal to -2 as follows:

```{r}
pt(-2, df = 5, lower.tail = FALSE)
```

This is the area under the $t$ distribution density curve to the right of -2.

```{r, echo = FALSE, message=FALSE, cache = TRUE, fig.height=2}
x_grid_1 <- seq(from = -2, to = 4, length = 101)

region_to_shade1 <- data.frame(
  x = c(-2, x_grid_1, 4),
  y = c(0, dt(x_grid_1, df = 14, log = FALSE), 0)
)

ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) +
  stat_function(fun = dt, args = list(df = 14)) +
  geom_polygon(
    mapping = aes(x = x, y = y),
    fill = "cornflowerblue",
    data = region_to_shade1) +
  geom_vline(xintercept = -2) +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.4), expand = FALSE) +
  theme_bw() +
  xlab("t") +
  ylab("t distribution density") +
  ggtitle(expression(paste("Shaded area is ", P(t >= -2))))
```

\newpage

The calculation of p-values depends on the value of the test statistic and the form of the alternative hypothesis.

```{r, echo = FALSE}
make_plot <- function(t, alt) {
  if(alt == ">") {
    x_grid_1 <- seq(from = t, to = 4, length = 101)

    region_to_shade1 <- data.frame(
      x = c(t, x_grid_1, 4),
      y = c(0, dt(x_grid_1, df = 14, log = FALSE), 0)
    )
    
    ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) +
      stat_function(fun = dt, args = list(df = 14)) +
      geom_polygon(
        mapping = aes(x = x, y = y),
        fill = "cornflowerblue",
        data = region_to_shade1) +
      geom_vline(xintercept = t) +
      coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.4), expand = FALSE) +
      theme_bw() +
      xlab("t") +
      ylab("")
  } else if(alt == "<") {
    x_grid_1 <- seq(from = -4, to = t, length = 101)

    region_to_shade1 <- data.frame(
      x = c(-4, x_grid_1, t),
      y = c(0, dt(x_grid_1, df = 14, log = FALSE), 0)
    )
    
    ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) +
      stat_function(fun = dt, args = list(df = 14)) +
      geom_polygon(
        mapping = aes(x = x, y = y),
        fill = "cornflowerblue",
        data = region_to_shade1) +
      geom_vline(xintercept = t) +
      coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.4), expand = FALSE) +
      theme_bw() +
      xlab("t") +
      ylab("")
  } else if(alt == "neq") {
    x_grid_1 <- seq(from = -4, to = -1 * abs(t), length = 101)

    region_to_shade1 <- data.frame(
      x = c(-4, x_grid_1, -1 * abs(t)),
      y = c(0, dt(x_grid_1, df = 14, log = FALSE), 0)
    )

    x_grid_2 <- seq(from = abs(t), to = 4, length = 101)

    region_to_shade2 <- data.frame(
      x = c(abs(t), x_grid_2, 4),
      y = c(0, dt(x_grid_2, df = 14, log = FALSE), 0)
    )
    
    ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) +
      stat_function(fun = dt, args = list(df = 14)) +
      geom_polygon(
        mapping = aes(x = x, y = y),
        fill = "cornflowerblue",
        data = region_to_shade1) +
      geom_polygon(
        mapping = aes(x = x, y = y),
        fill = "cornflowerblue",
        data = region_to_shade2) +
      geom_vline(xintercept = t) +
      geom_vline(xintercept = -t) +
      coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.4), expand = FALSE) +
      theme_bw() +
      xlab("t") +
      ylab("")
  }
}
```

```{r, echo = FALSE, warning=FALSE, fig.height = 7.5, fig.width = 7.3}
grid.newpage()
layout <- grid.layout(nrow = 6, ncol = 4,
  heights = unit(c(2, rep_len(1, 5)), units = c("lines", rep_len("null", 5))),
  widths = unit(c(1, 0.5, 1.5, 1.5), units = rep_len("null", 4))
)
pushViewport(viewport(layout = layout))

grid.text("Alternative\nHypothesis",
  vp = viewport(layout.pos.row = 1, layout.pos.col = 1))

grid.text("t\nStatistic",
  vp = viewport(layout.pos.row = 1, layout.pos.col = 2))

grid.text("R Code",
  vp = viewport(layout.pos.row = 1, layout.pos.col = 3))

grid.text("Picture",
  vp = viewport(layout.pos.row = 1, layout.pos.col = 4))

## H_A: parameter < null
grid.text(expression(paste(H[A]: mu < mu^{null})),
  vp = viewport(layout.pos.row = 2, layout.pos.col = 1))

grid.text("-2",
  vp = viewport(layout.pos.row = 2, layout.pos.col = 2),
  x = unit(0.5, "npc"),
  gp = gpar(fontsize = 11),
  just = c("center", "center"))

grid.text("pt(-2, df = 5)",
  vp = viewport(layout.pos.row = 2, layout.pos.col = 3),
  x = unit(0, "npc"),
  gp = gpar(fontsize = 11),
  just = c("left", "center"))

p <- make_plot(t = -2, alt = "<")
print(p,
  vp = viewport(layout.pos.row = 2, layout.pos.col = 4))



grid.text(expression(paste(H[A]: mu < mu^{null})),
  vp = viewport(layout.pos.row = 3, layout.pos.col = 1))

grid.text("2",
  vp = viewport(layout.pos.row = 3, layout.pos.col = 2),
  x = unit(0.5, "npc"),
  gp = gpar(fontsize = 11),
  just = c("center", "center"))

grid.text("pt(2, df = 5)",
  vp = viewport(layout.pos.row = 3, layout.pos.col = 3),
  x = unit(0, "npc"),
  gp = gpar(fontsize = 11),
  just = c("left", "center"))

p <- make_plot(t = 2, alt = "<")
print(p,
  vp = viewport(layout.pos.row = 3, layout.pos.col = 4))



## H_A: parameter > null
grid.text(expression(paste(H[A]: mu > mu^{null})),
  vp = viewport(layout.pos.row = 4, layout.pos.col = 1))

grid.text("-2",
  vp = viewport(layout.pos.row = 4, layout.pos.col = 2),
  x = unit(0.5, "npc"),
  gp = gpar(fontsize = 11),
  just = c("center", "center"))

grid.text("pt(-2, df = 5, lower.tail = FALSE)",
  vp = viewport(layout.pos.row = 4, layout.pos.col = 3),
  x = unit(0, "npc"),
  gp = gpar(fontsize = 11),
  just = c("left", "center"))

p <- make_plot(t = -2, alt = ">")
print(p,
  vp = viewport(layout.pos.row = 4, layout.pos.col = 4))



grid.text(expression(paste(H[A]: mu > mu^{null})),
  vp = viewport(layout.pos.row = 5, layout.pos.col = 1))

grid.text("2",
  vp = viewport(layout.pos.row = 5, layout.pos.col = 2),
  x = unit(0.5, "npc"),
  gp = gpar(fontsize = 11),
  just = c("center", "center"))

grid.text("pt(2, df = 5, lower.tail = FALSE)",
  vp = viewport(layout.pos.row = 5, layout.pos.col = 3),
  x = unit(0, "npc"),
  gp = gpar(fontsize = 11),
  just = c("left", "center"))

p <- make_plot(t = 2, alt = ">")
print(p,
  vp = viewport(layout.pos.row = 5, layout.pos.col = 4))



## H_A: parameter not equal to null
grid.text(expression(paste(H[A]: mu != mu^{null})),
  vp = viewport(layout.pos.row = 6, layout.pos.col = 1))

grid.text("-2 or 2",
  vp = viewport(layout.pos.row = 6, layout.pos.col = 2),
  x = unit(0.5, "npc"),
  gp = gpar(fontsize = 11),
  just = c("center", "center"))

grid.text("pt(-2, df = 5) +\n  pt(2, df = 5, lower.tail = FALSE)",
  vp = viewport(layout.pos.row = 6, layout.pos.col = 3),
  x = unit(0, "npc"),
  gp = gpar(fontsize = 11),
  just = c("left", "center"))

p <- make_plot(t = -2, alt = "neq")
print(p,
  vp = viewport(layout.pos.row = 6, layout.pos.col = 4))

```


