---
title: "Two Lines: Crocodiles!!"
output:
  pdf_document:
    keep_tex: true
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
require(readr)
```

## Multiple Regression

**ANOVA models have**:

 * a quantitative response variable (sepal width of a flower) and
 * one categorical explanatory variable (species)
 * Separate mean sepal width for each species, individual values normally distributed around the mean

**Simple linear regression models have**:

 * a quantitative response variable (college graduation rate) and 
 * one quantitative explanatory variable (college acceptance rate)
 * Mean graduation rate is a linear function of acceptance rate, individual values normally distributed around the mean

**Multiple regression models have**:

 * a quantitative response variable and
 * more than one explanatory variable, may be a mix of categorical and quantitative
 * Examples:
    * $\mu(Y|X_1, X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$
    * $\mu(Y|X_1, X_2, X_3, X_4) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4$
    * $\mu(Y|X_1, X_2) = \beta_0 + \beta_1 \log(X_1) + \beta_2 \log(X_2)$
    * $\mu(Y|X_1, X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 \times X_2 + \beta_4 X_1^2$

We will start by combining one categorical explanatory variable and one quantitative explanatory variable.

## Example of Two Lines

```{r, message = FALSE, echo = FALSE}
# set seed and sample_frac to get both species in the first few rows.
set.seed(123)

crocs <- read_csv("http://www.evanlray.com/data/sdm4/Crocodile_lengths.csv") %>%
  mutate(Species = factor(Species)) %>%
  sample_frac(size = 1) %>%
  as.data.frame()
```

We have measurements of the head length (cm) and total body length (cm) of 32 crocodiles of two different species:

```{r}
head(crocs)
nrow(crocs)
```

```{r, fig.height=2}
ggplot(data = crocs) +
  geom_point(mapping = aes(x = Body, y = Head, color = Species)) +
  theme_bw()
```

\newpage

## 2 lines by filtering to create separate data sets

```{r}
aus_crocs <- crocs %>% filter(Species == "Australian")
aus_fit <- lm(Head ~ Body, data = aus_crocs)
summary(aus_fit)
```

```{r}
ind_crocs <- crocs %>% filter(Species == "Indian")
ind_fit <- lm(Head ~ Body, data = ind_crocs)
summary(ind_fit)
```

#### Questions we'd like to be able to answer (but can't with this output):

1. How strong is the evidence that the intercepts for these lines are different?
2. How strong is the evidence that the slopes for these lines are different?

\newpage

## 2 parallel lines (same slope)

* Our Goal: Equations for two lines
\begin{align*}
\text{Estimated Mean Head Length for Australian Crocs } &= \hat{\beta}_0^{Australian} + \hat{\beta}_{1} \times (\text{Body Length}) \\
\text{Estimated Mean Head Length for Indian Crocs } &= \hat{\beta}_0^{Indian} + \hat{\beta}_{1} \times (\text{Body Length})
\end{align*}

 * Note: Different intercepts, same slope.

```{r, echo = TRUE, fig.height=2}
parallel_lines_fit <- lm(Head ~ Body + Species, data = crocs)
summary(parallel_lines_fit)

crocs <- crocs %>%
  mutate(
    fitted = predict(parallel_lines_fit)
  )

ggplot(data = crocs) +
  geom_point(mapping = aes(x = Body, y = Head, color = Species)) +
  geom_line(mapping = aes(x = Body, y = fitted, color = Species))
```

* R gives us a single combined equation:
$$\text{Estimated Mean Head Length} = \hat{\mu} = \hat{\beta}_0 + \hat{\beta}_{1} \text{Body} + \hat{\beta}_2 \text{SpeciesIndian}$$

$$\hat{\mu} = 2.27 + 0.13 \text{Body} + 8.89 \text{SpeciesIndian}$$

\newpage

#### What is the `SpeciesIndian` variable?

* Behind the scenes, R creates a new **indicator variable** called `SpeciesIndian`:

$$\text{{\tt SpeciesIndian}} = \begin{cases} 1 & \text{ if the species for crocodile $i$ is Indian.} \\
0 & \text{ otherwise (in this case, the species is Australian)} \end{cases}
$$

* R doesn't modify the data frame (it creates a secret copy in the background), but it would look like this:

```{r, echo = FALSE}
crocs <- mutate(crocs,
  SpeciesIndian = as.integer(Species == "Indian"))
```

```{r, echo = TRUE}
head(crocs)
```

Above, we obtained this estimated equation:

$$\hat{\mu} = 2.27 + 0.13 (\text{Body}) + 8.89 \text{SpeciesIndian}$$

#### What is the estimated equation describing the relationship between body length and head length, for Australian crocodiles?

\vspace{2cm}

#### What is the estimated equation describing the relationship between body length and head length, for Indian crocodiles?

\vspace{2cm}

#### What is the interpretation of $\widehat{\beta}_0 = 2.27$?

\vspace{2cm}

#### What is the interpretation of $\widehat{\beta}_1 = 0.13$?

\vspace{2cm}

#### What is the interpretation of $\widehat{\beta}_2 = 8.89$?

\newpage

#### Using the output from the summary function, conduct a test of the claim that a single regression line can be used to describe the relationship between body length and head length in the population of all Australian and Indian crocodiles.

\vspace{3cm}

#### Conduct a test of the claim that neither species nor body length are associated with head length in the population of all Australian and Indian crocodiles.  (Note: formally, this is a test only of linear association with body length.)

\vspace{3cm}

#### Find and interpret a 95% confidence interval for $\beta_2$, the coefficient of `SpeciesIndian`.

```{r}
confint(parallel_lines_fit)
```

\vspace{3cm}

#### Find and interpret a 95% confidence interval for the mean head length of the sub-population of Australian crocodiles that have a total body length of 400cm.

```{r}
predict_data <- data.frame(
  Species = "Australian",
  Body = 400
)

predict(parallel_lines_fit, newdata = predict_data, interval = "confidence")
```

\newpage

## 1 model, 2 lines (different slopes)

 * Our Goal: Equations for two lines
\begin{align*}
\text{Estimated Mean Head Length for Australian Crocs } &= \hat{\beta}_0^{Australian} + \hat{\beta}_{1}^{Australian} \times (\text{Body Length}) \\
\text{Estimated Mean Head Length for Indian Crocs } &= \hat{\beta}_0^{Indian} + \hat{\beta}_{1}^{Indian} \times (\text{Body Length})
\end{align*}

 * Note: Different intercepts and slopes.

 * To allow for different slopes, you have two options for the formula (the model that it fits is the same either way, the second is just a shorthand for the first):
    * `Body + Species + Body:Species`
    * `Body * Species`

 * The term `Body:Species` is called the **interaction** between `Body` and `Species`.  It is just the product of those two variables.

```{r, echo = TRUE, fig.height=2}
two_lines_fit <- lm(Head ~ Body + Species + Body:Species, data = crocs)
summary(two_lines_fit)

crocs <- crocs %>%
  mutate(
    fitted = predict(two_lines_fit)
  )

ggplot(data = crocs) +
  geom_point(mapping = aes(x = Body, y = Head, color = Species)) +
  geom_line(mapping = aes(x = Body, y = fitted, color = Species))
```

#### What is the estimated equation for the mean from this model?

\vspace{2.5cm}

#### What is the estimated equation describing the relationship between body length and head length, for Australian crocodiles?

\vspace{2.5cm}

#### What is the estimated equation describing the relationship between body length and head length, for Indian crocodiles?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_0 = 3.463$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_1 = 7.075$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_2 = 0.125$?

\vspace{2.5cm}

#### What is the interpretation of $\widehat{\beta}_3 = 0.006$?

\newpage

#### Using the output from the summary function, conduct a test of the claim that the *slope* of the line describing the relationship between body length and head length in the population of all Australian crocodiles is the same as the *slope* of the line describing the relationship between body length and head length in the population of all Indian crocodiles.

