---
title: ""
subtitle: ""
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    keep_tex: true
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

## Stat 242 Quiz -- Topics Drawn from Chapters 9 and 10

## *R Output Appendix*

The HELP study was a clinical trial for adult inpatients recruited from a detoxification unit. Patients with no primary care physician were randomized to receive a multidisciplinary assessment and a brief motivational intervention or usual care, with the goal of linking them to primary medical care.  As part of the study, a number of measurements were taken on the participants at baselne.  Here we will look at the following three variables:

 * `substance` is the primary substance of abuse: a variable with three levels (`alcohol`, `cocaine`, and `heroin`)
 * `mcs` is the SF-36 Mental Component Score (measured at baseline, lower scores indicate worse mental health status)
 * `i1` is the average number of drinks (standard units) consumed per day, in the past 30 days (measured at baseline)

We will use `i1` as the response and `substance` and `mcs` as explanatory variables.

```{r, echo=FALSE, message=FALSE}
library(mosaicData)
library(ggplot2)
library(dplyr)
library(gridExtra)
```

## Initial plots

```{r, fig.height = 3, fig.width = 7.5}
p1 <- ggplot(data = HELPrct, mapping = aes(x = mcs, y = i1, color = substance)) +
  geom_smooth(method = "lm") +
  geom_point()

p2 <- ggplot(data = HELPrct, mapping = aes(x = i1, color = substance)) +
  geom_density()

grid.arrange(p1, p2, ncol = 2)
```

\newpage

## Plots with transformed response variable

```{r, echo = TRUE}
HELPrct <- HELPrct %>%
  mutate(sqrt_i1 = sqrt(i1))

ggplot(data = HELPrct, mapping = aes(x = mcs, y = sqrt_i1, color = substance)) +
  geom_smooth(method = "lm") +
  geom_point()

lm_fit <- lm(sqrt_i1 ~ substance + mcs, data = HELPrct)

HELPrct <- HELPrct %>%
  mutate(
    residual = residuals(lm_fit)
  )

ggplot(data = HELPrct, mapping = aes(x = mcs, y = residual)) +
  geom_point()

ggplot(data = HELPrct, mapping = aes(x = residual, color = substance)) + 
  geom_density()
```

## Summary Output -- Model 1

```{r}
lm_fit1 <- lm(sqrt_i1 ~ substance + mcs, data = HELPrct)
summary(lm_fit1)
confint(lm_fit1)
```

Define the following notation for the coefficients in Model 1:

* $\beta_0^{M1}$ labeled `(Intercept)`
* $\beta_1^{M1}$ labeled `substancecocaine`
* $\beta_2^{M1}$ labeled `substanceheroin`
* $\beta_3^{M1}$ labeled `mcs`

\newpage

## Summary Output -- Model 2

```{r}
lm_fit2 <- lm(sqrt_i1 ~ substance * mcs, data = HELPrct)
summary(lm_fit2)
confint(lm_fit2)
```


Define the following notation for the coefficients in Model 2:

* $\beta_0^{M2}$ labeled `(Intercept)`
* $\beta_1^{M2}$ labeled `substancecocaine`
* $\beta_2^{M2}$ labeled `substanceheroin`
* $\beta_3^{M2}$ labeled `mcs`
* $\beta_4^{M2}$ labeled `substancecocaine:mcs`
* $\beta_5^{M2}$ labeled `substanceheroin:mcs`

## ANOVA comparison of Model 1 and Model 2

```{r}
anova(lm_fit1, lm_fit2)
```
