---
title: "ANOVA: Indicator Variables Formulation"
subtitle: "Sleuth3 Sections 5.2, 6.2, 9.3.2 and 9.3.3"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    keep_tex: true
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
options(width = 100)
```

# Iris Data Again

## Example 1: Sepal Width of Iris Flowers

### Study Overview

We have measurements of the characteristics of 150 iris flowers, 50 each of three different species:

 * Iris setosa is found in the arctic, including Alaska and Maine in the United States, Canada, Russia, northern China, Korea and other northern countries.
 * Iris versicolor is found in the eastern United States and eastern Canada.
 * Iris virginica is found in the eastern United States

### Look at the Data:

```{r, message = FALSE, echo = FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
library(gmodels)

options("pillar.sigfig" = 10) # print 10 significant digits in summarize output
```

```{r}
head(iris)
```

```{r}
# Calculate sample means and standard deviations separately for each species
iris %>%
  group_by(Species) %>%
  summarize(
    mean = mean(Sepal.Width),
    sd = sd(Sepal.Width)
  )
```

### Parameters:

$\mu_1$ = Average sepal width among all setosa flowers (in the region where the flowers in the sample were found?)

$\mu_2$ = Average sepal width among all versicolor flowers (in the region where the flowers in the sample were found?)

$\mu_3$ = Average sepal width among all virginica flowers (in the region where the flowers in the sample were found?)

\newpage

### Indicator variable parameterization

R's output (and output from most other statistical packages) directly answers 3 questions:

1. What is an estimate of $\mu_1$?
2. What is an estimate of $\mu_2 - \mu_1$?
3. What is an estimate of $\mu_3 - \mu_1$?

Verification:

```{r}
anova_fit <- lm(Sepal.Width ~ Species, data = iris)
summary(anova_fit)
```

#### 1. Compare the `Estimate` labeled `(Intercept)` to:

* the group mean for setosa flowers: 3.428

#### 2. Compare the `Estimate` labeled `Speciesversicolor` to:

 * the difference in group means for setosa and versicolor flowers: 2.770 - 3.428 = -0.658
 * the results of a hypothesis test that $\mu_2 - \mu_1 = 0$:
 
```{r}
fit.contrast(anova_fit, "Species", c(-1, 1, 0), conf = 0.95)
```

#### 3. Compare the `Estimate` labeled `Speciesvirginica` to:

 * the difference in group means for setosa and virginica flowers: 2.974 - 3.428 = -0.454
 * the results of a hypothesis test that $\mu_3 - \mu_1 = 0$:
 
```{r}
model_fit <- lm(Sepal.Width ~ Species, data = iris)
fit.contrast(model_fit, "Species", c(-1, 0, 1), conf = 0.95)
```

\newpage

### Getting to estimates of means from the R output

We express the mean for a flower of a particular species as follows:

$\mu = \beta_0 + \beta_1 Speciesversicolor + \beta_2 Speciesvirginica$

Here, $Speciesversicolor$ and $Speciesvirginica$ are new **indicator variables**:

 * $Speciesversicolor = \begin{cases} 1 \text{ if a flower is of the versicolor species} \\ 0 \text{ otherwise} \end{cases}$

 * $Speciesvirginica = \begin{cases} 1 \text{ if a flower is of the virginica species} \\ 0 \text{ otherwise} \end{cases}$

In the background, R creates a new copy of our data frame with these indicator variables that looks like this:

```{r, echo = FALSE}
iris_augmented <- iris %>%
  mutate(
    Speciesversicolor = as.numeric(Species == "versicolor"),
    Speciesvirginica = as.numeric(Species == "virginica")
  )
```

```{r, echo = TRUE, eval=FALSE}
iris_augmented
```

```{r, echo = FALSE, eval = FALSE}
iris_augmented[1:10, ]
```

\begin{verbatim}
##    Sepal.Length Sepal.Width Petal.Length Petal.Width    Species Speciesversicolor Speciesvirginica
## 1           5.1         3.5          1.4         0.2     setosa                 0                0
## 2           4.9         3.0          1.4         0.2     setosa                 0                0
## 3           4.7         3.2          1.3         0.2     setosa                 0                0
\end{verbatim}

\[\vdots\]

\begin{verbatim}
## 48          4.6         3.2          1.4         0.2     setosa                 0                0
## 49          5.3         3.7          1.5         0.2     setosa                 0                0
## 50          5.0         3.3          1.4         0.2     setosa                 0                0
## 51          7.0         3.2          4.7         1.4 versicolor                 1                0
## 52          6.4         3.2          4.5         1.5 versicolor                 1                0
## 53          6.9         3.1          4.9         1.5 versicolor                 1                0
\end{verbatim}

\[\vdots\]

\begin{verbatim}
## 98           6.2         2.9          4.3         1.3 versicolor                1                0
## 99           5.1         2.5          3.0         1.1 versicolor                1                0
## 100          5.7         2.8          4.1         1.3 versicolor                1                0
## 101          6.3         3.3          6.0         2.5  virginica                0                1
## 102          5.8         2.7          5.1         1.9  virginica                0                1
## 103          7.1         3.0          5.9         2.1  virginica                0                1
\end{verbatim}

#### Express the mean for *setosa* flowers in terms of $\beta_0$, $\beta_1$, and/or $\beta_2$

\vspace{1cm}

#### Express the mean for *versicolor* flowers in terms of $\beta_0$, $\beta_1$, and/or $\beta_2$

\vspace{1cm}

#### Express the mean for *virginica* flowers in terms of $\beta_0$, $\beta_1$, and/or $\beta_2$

\vspace{1cm}


In the output from `summary(anova_fit)`, we have:

 * The row labeled `(Intercept)` is related to $\beta_0$ (estimate and test of $H_0: \beta_0 = 0$)
 * The row labeled `Speciesversicolor` is related to $\beta_1$ (estimate and test of $H_0: \beta_1 = 0$)
 * The row labeled `Speciesvirginica` is related to $\beta_2$ (estimate and test of $H_0: \beta_2 = 0$)

\newpage

### Changing the baseline category

Here's a thing you don't really need to know how to do; just showing that it is possible.

Suppose that instead of using `setosa` for the baseline species, we want to use `virginica` as the baseline.

```{r}
# What are the levels of the Species variable in the iris data frame, in order?
levels(iris$Species)

# Update the levels to be in the order "virginica" first, "versicolor" second, and "setosa" third
iris <- iris %>%
  mutate(
    Species = factor(Species, levels = c("virginica", "versicolor", "setosa"))
  )

# Fit with updated order of levels
fit2 <- lm(Sepal.Width ~ Species, data = iris)
summary(fit2)
```

\newpage

### For you to do:

Back in Lab 2 on RStudio, look at the output from calling `summary` on your linear model fit object.  Answer the questions below.

 * What is the baseline category for the explanatory variable in this model?
 
\vspace{3cm}

 * What are the possible values of the `JudgeSpock's` variable, and in what circumstances does the variable equal each of those values?

\vspace{3cm}

 * What is the `Estimate` labelled `(Intercept)` an estimate of?  Be as precise as possible.

\vspace{3cm}

 * What is the `Estimate` labelled `JudgeSpock's` an estimate of?  Be as precise as possible.

\vspace{3cm}

 * Use the output from `summary` conduct a test of the null hypothesis that $\mu_A = \mu_{Spock's}$.

\vspace{3cm}

 * Could you use the output from `summary` conduct a test of the null hypothesis that $\mu_B = \mu_{Spock's}$?  (The answer is no - why not?)

