---
title: "ANOVA: Conditions"
subtitle: "Sleuth3 Chapter 3 and Section 5.5"
date: "20190930"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## Summary

#### Model Statement:

 * Observations are independent of each other, and observations in group $i$ follow a Normal($\mu_i$, $\sigma^2$) distribution

```{r, echo = FALSE, fig.height = 1.3, message=FALSE}
library(ggplot2)
library(dplyr)
library(readr)

x_grid <- seq(from = 0, to = 20, length = 101)
group_sd <- 2

plot_df <- data.frame(
  x = rep(x_grid, 3),
  y = c(dnorm(x_grid, mean = 8, sd = group_sd),
        dnorm(x_grid, mean = 10, sd = group_sd),
        dnorm(x_grid, mean = 14, sd = group_sd)),
  group = rep(c("Group 1: mean = 8, sd = 2", "Group 2: mean = 10, sd = 2", "Group 3: mean = 14, sd = 2"),
    each = length(x_grid))
)

ggplot(data = plot_df, mapping = aes(x = x, y = y, color = group)) +
  geom_line()
  
```

#### Conditions to check, relative importance, and what to do if not met

 * Observations are **independent** (knowing that one observation is above its group mean wouldn't give you any information about whether or not another observation is above its group mean)
    * Very important
    * Suspect if data are collected over time or we have multiple observations for very similar people (twins, baseline and post-treatment, ...)
    * Use a different model
 * **Normal** distribution
    * Not that important, especially if $n$ is large
    * Try a transformation; permutation or bootstrap methods for inference; don't worry too much
 * **Equal variance** for all groups
    * Very important
    * Transform the data; permutation or bootstrap methods for inference; or use a different model
 * **No outliers** (not a formal part of the model, but important to check in practice)
    * Potentially important; do they affect the results?
    * Try a transformation
    * Run analysis both with and without outliers; REPORT BOTH ANALYSES

In the next couple of days we will focus on data transformations.

We saw permutation tests in the first week of class and will review them again later this week.  We will likely cover bootstrap methods next week.

## Effects of condition violations

 * Generally, estimates of group means are OK (unless outliers are severe and $n$ is small)
 * Confidence intervals and p-values based on $t$ and $F$ distributions could be more affected

For hypothesis tests:

 * If conditions are met, the p-value accurately describes the probability of obtaining a test statistic at least as extreme as the statistic we observed, if the null hypothesis is true.
 * If not, the probability of obtaining a test statistic at least as extreme as the statistic we observed may be higher or lower than the reported p-value.

For confidence intervals:

 * If conditions are met, for 95% of samples a 95% CI based on that sample will contain the parameter being estimated.
 * If not, 95% CIs may contain the parameter for more or less than 95% of samples.

\newpage

#### Non-normal distributions (Figure 3.4 in Sleuth3)

\includegraphics[width=0.85\textwidth]{coverage_2sample_nonnormal.png}

#### Non-equal variances (Figure 3.5 in Sleuth3)

\includegraphics[width=0.85\textwidth]{coverage_2sample_nonequal_var.png}

\newpage

#### What do normally distributed data look like?

 * You should not expect a perfect bell curve, especially if your sample size is small
 * Here are histograms and density plots of 30 different samples of size 15 from a Normal(0, 1) distribution:

```{r, echo = FALSE, fig.height = 9}
set.seed(13245)
fake_data <- data.frame(
  x = rnorm(15*30),
  sample_ind = factor(rep(1:30, each = 15))
)

ggplot(data = fake_data, mapping = aes(x = x)) +
  geom_histogram(bins = 30, mapping = aes(y = ..density..)) +
  geom_density() +
  facet_wrap( ~ sample_ind, ncol = 3) +
  theme_bw()
```

\newpage

## Example Data Sets

#### Irises

Here are density plots and box plots, separately for each Species.

```{r, fig.height = 2.5}
ggplot(data = iris, mapping = aes(x = Sepal.Width, color = Species)) +
  geom_density()

ggplot(data = iris, mapping = aes(x = Species, y = Sepal.Width)) +
  geom_boxplot()
```

Standard deviations for each group:

```{r}
iris %>%
  group_by(Species) %>%
  summarize(
    sd_sepal_width = sd(Sepal.Width)
  )
```

\newpage

Here are plots of residuals vs. group means for each group, as well as the standard deviations within each group:

```{r}
species_fit <- lm(Sepal.Width ~ Species, data = iris)
iris <- iris %>%
  mutate(
    fitted = fitted(species_fit),
    residual = residuals(species_fit)
  )

ggplot(data = iris, mapping = aes(x = fitted, y = residual, color = Species)) +
  geom_point()
```

 * **Independent**?

\vspace{2cm}

 * **Normal distribution** within each group?

\vspace{2cm}

 * **Equal variance** for all groups?  (approximately)

\vspace{2cm}

 * **Outliers**?

\vspace{2cm}

\newpage

#### Cloud Seeding (Sleuth3 Case Study 3.1.1)

Quote from book: "On each of 52 days that were deemed suitable for cloud seeding, a random mechanism was used to decide whether to seed the target cloud on that day or to leave it unseeded as a control.  An airplane flew through the cloud in both cases.... [p]recipitation was measured as the total rain volume falling from the cloud base following the airplane seeding run."

```{r, message = FALSE}
clouds <- read_csv("http://www.evanlray.com/data/sleuth3/case0301_cloud_seeding.csv")
head(clouds)
```

Here are density plots and box plots, separately for each Treatment.

```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = Rainfall, color = Treatment)) +
  geom_density()

ggplot(data = clouds, mapping = aes(x = Treatment, y = Rainfall)) +
  geom_boxplot()
```

Standard deviations for each group:

```{r}
clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(Rainfall)
  )
```

Here is a plot of residuals vs. fitted/predicted responses for each group:

```{r, fig.height=2.5}
clouds_fit <- lm(Rainfall ~ Treatment, data = clouds)
clouds <- clouds %>%
  mutate(
    fitted = fitted(clouds_fit),
    residual = residuals(clouds_fit)
  )

ggplot(data = clouds, mapping = aes(x = fitted, y = residual, color = Treatment)) +
  geom_point()
```

 * **Independent**?

\vspace{2cm}

 * **Normal distribution** within each group?

\vspace{2cm}

 * **Equal variance** for all groups?  (approximately)

\vspace{2cm}

 * **Outliers**?

\vspace{2cm}

\newpage

#### Solar Radiation and Skin Cancer (Sleuth3 Exercise 3.23)

We have data on yearly skin cancer rates (cases per 100,000 people) in Connecticut from 1938 to 1972.  We also have recorded whether each year came 2 years after high than average sunspot activity, or 2 years after lower than average sunspot activity.

```{r, message = FALSE}
cancer <- read_csv("http://www.evanlray.com/data/sleuth3/ex0323_skin_cancer.csv")
head(cancer)
```

Here are density plots and box plots, separately for each level of SunspotActivity

```{r, fig.height = 1.9}
ggplot(data = cancer, mapping = aes(x = CancerRate, color = SunspotActivity)) +
  geom_density()

ggplot(data = cancer, mapping = aes(x = SunspotActivity, y = CancerRate)) +
  geom_boxplot()
```

Standard deviations for each group:

```{r}
cancer %>%
  group_by(SunspotActivity) %>%
  summarize(
    sd_rainfall = sd(CancerRate)
  )
```

\newpage

Here is a plot of residuals vs. fitted/predicted responses for each group:

```{r, fig.height = 1.9}
cancer_fit <- lm(CancerRate ~ SunspotActivity, data = cancer)
cancer <- cancer %>%
  mutate(
    fitted = fitted(cancer_fit),
    residual = residuals(cancer_fit)
  )

ggplot(data = cancer, mapping = aes(x = fitted, y = residual, color = SunspotActivity)) +
  geom_point()
```

When time is involved, it can be informative to plot the residuals vs time:

```{r, fig.height = 1.9}
ggplot(data = cancer, mapping = aes(x = Year, y = residual)) +
  geom_point() +
  facet_wrap( ~ SunspotActivity)
```


 * **Independent**?

\vspace{1.9cm}

 * **Normal distribution** within each group?

\vspace{1.9cm}

 * **Equal variance** for all groups?  (approximately)

\vspace{1.9cm}

 * **Outliers**?

\vspace{1.9cm}




