---
title: "Concepts: Transformations for ANOVA models"
subtitle: "20190930 -- Sleuth3 Sections 3.5 and 5.5"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(gridExtra)
library(ggplot2)
library(dplyr)
```

## Context

 * Transformations can sometimes help with the following issues:
    * non-normal distributions within each group (but skewness is only a problem if it is very serious)
    * lack of equal variance for all groups
    * outliers (but usually only if this is a side effect of serious skewness)
 * The most common transformations (that we'll consider in this class) work for positive numbers only.

## The Ladder of Powers

 * Imagine a "ladder of powers" of $y$ (or $x$): We start at $y$ and go up or down the ladder.

\begin{table}[!h]
\centering
\begin{tabular}{c c p{8cm}}
\toprule
Transformation & R Code & Comments \\
\toprule
\vdots & \\
\midrule
$e^y$ & \verb&exp(y)& & Exactly where on the ladder the exponential transformation belongs depends on the magnitude of the data, but somewhere around here... \\
\midrule
$y^2$ & \verb&y^2& & \\
\midrule
$y$ &  & Start here (no transformation) \\
\midrule
$\sqrt{y}$ & \verb&sqrt(y)& & \\
\midrule
$y^{``0"}$ & \verb&log(y)& & We use $\log(y)$ here \\
\midrule
$-1/\sqrt{y}$ & \verb&-1/sqrt(y)& & The $-$ keeps the values of $y$ in order \\
\midrule
$-1/y$ & \verb&-1/y& & \\
\midrule
$-1/y^2$ & \verb&-1/y^2& & \\
\midrule
\vdots & \\
\bottomrule
\end{tabular}
\end{table}

## Some (minimal) facts about logarithms and exponentials

 * Foundations:
    * In this class the base of our logarithms is $e$
    * Notation: $\exp(x) = e^x$
 * $\log()$ and $\exp()$ are inverses
    * $\log(\exp(x)) = x$
    * $\exp(\log(x)) = x$
 * They are useful because they convert multiplication to addition, and addition to multiplication
    * $\log(a \cdot b) = \log(a) + \log(b)$
    * $\exp(a + b) = \exp(a) \cdot \exp(b)$

\newpage

 * Which direction?
    * If a variable is skewed right, move it down the ladder (pull down large values)
    * If a variable is skewed left, move it up the ladder (pull up small values)

```{r, echo = FALSE}
example <- data.frame(
  y = c(1, 2, 3, 4),
  y_squared = c(1, 2, 3, 4)^2,
  y_cubed = c(1, 2, 3, 4)^3,
  sqrt_y = c(1, 2, 3, 4)^0.5,
  log_y = log(c(1, 2, 3, 4))
)
```

```{r, fig.height = 5, echo = FALSE}
p0 <- ggplot(data = example, mapping = aes(x = y_cubed, y = 0)) +
  geom_point() +
  ggtitle(expression(paste("2 Steps Up from Goal: ", y^3, " is very skewed right"))) +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

p1 <- ggplot(data = example, mapping = aes(x = y_squared, y = 0)) +
  geom_point() +
  ggtitle(expression(paste("1 Step Up from Goal: ", y^2, " is slightly skewed right"))) +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

p2 <- ggplot(data = example, mapping = aes(x = y, y = 0)) +
  geom_point() +
  ggtitle("Goal: y is symmetric") +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

p3 <- ggplot(data = example, mapping = aes(x = sqrt_y, y = 0)) +
  geom_point() +
  ggtitle(expression(paste("1 Step Down from Goal: ", sqrt(y), " is slightly skewed left"))) +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

p4 <- ggplot(data = example, mapping = aes(x = log_y, y = 0)) +
  geom_point() +
  ggtitle("2 Steps Down from Goal: log(y) is very skewed left") +
  ylab("") +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

grid.arrange(p0, p1, p2, p3, p4, ncol = 1)
```

\newpage

\includegraphics{transformations.pdf}
