---
title: "Examples: Transformations for ANOVA models"
subtitle: "20190930 -- Sleuth3 Sections 3.5 and 5.5"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    highlight: zenburn
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(gridExtra)
library(readr)
library(dplyr)
library(ggplot2)
library(gmodels)
```

## Example: Cloud Seeding (Sleuth3 Case Study 3.1.1)

Quote from book: "On each of 52 days that were deemed suitable for cloud seeding, a random mechanism was used to decide whether to seed the target cloud on that day or to leave it unseeded as a control. ... [P]recipitation was measured as the total rain volume falling from the cloud base following the airplane seeding run."

```{r, message = FALSE}
clouds <- read_csv("http://www.evanlray.com/data/sleuth3/case0301_cloud_seeding.csv")
head(clouds, 4)
```

#### Starting Point

Here are density plots and box plots, separately for each Treatment.

```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = Rainfall, color = Treatment)) +
  geom_density()
```

Standard deviations for each group:

```{r}
clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(Rainfall)
  )
```

The standard deviations are very different, and the distributions are skewed right, so move down one step on the ladder.

\newpage

#### Down 1 Step: $\sqrt{Rainfall}$

```{r}
clouds <- clouds %>%
  mutate(
    sqrt_rainfall = sqrt(Rainfall)
  )
```


```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = sqrt_rainfall, color = Treatment)) +
  geom_density()

clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(sqrt_rainfall)
  )
```

These distributions are closer to symmetric -- probably good enough.

The ratio of these standard deviations is less than 2 -- often used as a guide for when we're OK.

However, we can make it even better if we go down another step.

\newpage

#### Down 2 Steps: $\log(Rainfall)$

```{r}
clouds <- clouds %>%
  mutate(
    log_rainfall = log(Rainfall)
  )
```

```{r, fig.height = 2.9}
ggplot(data = clouds, mapping = aes(x = log_rainfall, color = Treatment)) +
  geom_density()

clouds %>%
  group_by(Treatment) %>%
  summarize(
    sd_rainfall = sd(log_rainfall)
  )
```

Good enough!  We can conduct our analysis on this scale.

\newpage

#### Analysis on transformed scale

```{r}
clouds %>%
  group_by(Treatment) %>%
  summarize(
    mean_log_rainfall = mean(log_rainfall)
  )

rainfall_fit <- lm(log_rainfall ~ Treatment, data = clouds)
summary(rainfall_fit)
confint(rainfall_fit)

library(gmodels)
fit.contrast(rainfall_fit, "Treatment", c(1, -1), conf.int = 0.95)
```

We can interpret these numbers either on the new, transformed, data scale or on the original data scale.

\newpage

#### 1. Interpret the group mean estimates above on the transformed scale (always works!):

\vspace{6cm}

#### 2. Interpret the group mean estimates above on the original data scale (works if we got to a place where distributions were approximately symmetric after transformation!):

```{r}
exp(5.13)
exp(3.99)
```

\newpage

#### 3. Interpret the estimated difference in means above on the transformed scale (always works!):

\vspace{6cm}

#### 4. Interpret the estimted difference in means above on the original data scale (works only if the transformation selected was the log transformation and the resulting distribution was approximately symmetric!):

```{r}
exp(1.143781)
exp(0.240865)
exp(2.046697)
```

