---
title: "Quadratic Regression"
subtitle: "Oct. 25 2019"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
  - \usepackage{vwcol}
geometry: margin=0.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
library(ggplot2)
library(readr)
library(dplyr)
library(gridExtra)
options(width = 100)
```

#### Summary

 * Sometimes there is actually a quadratic relationship between two variables: $\mu(Y | X) = \beta_0 + \beta_1 X + \beta_2 X^2$
 * Fit with `lm(response ~ poly(explanatory, degree = 2, raw = TRUE), data = data)`

#### Galileo's data on falling things (Sleuth3 Case study 10.1.1)

 * Galileo showed that the trajectory of a body falling with horizontal velocity is a parabola.
 * Rolled ink-covered bronze ball down an inclined plane
 * Measured vertical height and horizontal distance in units of punti (1 punto is 169/180 mm)

\includegraphics{sleuth3_display_10_1.png}

```{r, echo = FALSE}
galileo <- Sleuth3::case1001
```

```{r, fig.height = 3, fig.width=4}
ggplot(data = galileo, mapping = aes(x = Height, y = Distance)) +
  geom_point() +
  theme_bw()
```

\newpage

```{r}
quadratic_fit <- lm(Distance ~ poly(Height, degree = 2, raw = TRUE), data = galileo)
summary(quadratic_fit)
```

```{r, fig.height = 3, fig.width=4}
ggplot(data = galileo, mapping = aes(x = Height, y = Distance)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, degree = 2, raw = TRUE), se = FALSE) +
  theme_bw()
```

#### 1. What model did we fit?

\vspace{2cm}

#### 2. What is the estimated equation describing the relationship between height and "population" mean distance?

