---
title: ""
subtitle: ""
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    keep_tex: true
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

## Stat 242 Quiz -- Topics Drawn from Chapters 9 through 12

## What's Your Name? ____________________

```{r, echo = FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
library(gridExtra)
library(car)

mammals <- Sleuth3::case0902 %>%
  select(
    Species, Body, Gestation, Litter, Brain
  )
```

We have a data set with the following information about different species of mammals:

 * `Species`: The species of mammal
 * `Body`: Average weight of the body
 * `Gestation`: Average length of pregnancy
 * `Litter`: Average litter size
 * `Brain`: Average weight of the brain

We will use brain size as the response variable and the other variables as explanatory variables.  Here is a look at the first few rows of the data, as well as the species in the data set. Note that there appear to be some closely related species in the data set; for example, there are three species of Porcupine, and four species of Deer mouse.

```{r}
head(mammals)
mammals$Species
```

### Initial Set Up

```{r, fig.height = 3.5}
ggpairs(mammals %>% select(-Species))
```

```{r, fig.height = 3.5}
mammals_transformed <- mammals %>%
  mutate(
    Body = -1/(Body^0.1),
    Gestation = -1/(Gestation^0.1),
    Litter = -1/(Litter^0.1),
    Brain = -1/(Brain^0.1)
  )

ggpairs(mammals_transformed %>% select(-Species))
```

\newpage

### Model 1: All Observations

```{r}
lm_fit <- lm(Brain ~ Body + Gestation + Litter, data = mammals_transformed)
summary(lm_fit)
confint(lm_fit)
```

#### Examining Residuals

Let's look at the residuals plots.  Recalling that there were three species of Porcupines in the data set, I have used a different color for the residuals for those species.

```{r, fig.height = 3.5}
mammals_transformed <- mammals_transformed %>%
  mutate(
    is_Porcupine = Species %in% c("Porcupine I", "Porcupine II", "Porcupine III"),
    residual = residuals(lm_fit)
  )

p1 <- ggplot(data = mammals_transformed, mapping = aes(x = Body, y = residual, color = is_Porcupine)) +
  geom_point()
p2 <- ggplot(data = mammals_transformed, mapping = aes(x = Gestation, y = residual, color = is_Porcupine)) +
  geom_point()
p3 <- ggplot(data = mammals_transformed, mapping = aes(x = Litter, y = residual, color = is_Porcupine)) +
  geom_point()
p4 <- ggplot(data = mammals_transformed, mapping = aes(x = residual)) +
  geom_density()

grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

Here are the residuals for the three species of porcupines in the data set:

```{r}
mammals_transformed %>%
  filter(is_Porcupine) %>%
  pull(residual)
```

#### Variance Inflation Factor and added variable plot

```{r, fig.height = 3.5}
vif(lm_fit)
avPlots(lm_fit)
```

\newpage

### Model 2: Setting aside some observations

```{r, fig.height = 3.5}
mammals_transformed <- mammals_transformed %>%
  mutate(
    obs_index = row_number(),
    h = hatvalues(lm_fit),
    studres = rstudent(lm_fit),
    D = cooks.distance(lm_fit)
  )

p1 <- ggplot(data = mammals_transformed, mapping = aes(x = obs_index, y = h)) +
  geom_hline(yintercept = 2*4/nrow(mammals_transformed))+
  geom_point()

p2 <- ggplot(data = mammals_transformed, mapping = aes(x = obs_index, y = studres)) +
  geom_point()

p3 <- ggplot(data = mammals_transformed, mapping = aes(x = obs_index, y = D)) +
  geom_point()

grid.arrange(p1, p2, p3, ncol = 1)
```

```{r}
obs_to_investigate <- c(28, 38, 53, 64, 72)

mammals_transformed <- mammals_transformed %>%
  mutate(
    suspicious = row_number() %in% obs_to_investigate
  )
mammals_no_suspicious <- mammals_transformed %>% filter(!suspicious)
```

\newpage

```{r}
lm_fit_no_suspicious <- lm(Brain ~ Body + Gestation + Litter, data = mammals_no_suspicious)
summary(lm_fit_no_suspicious)
```

\newpage

## Problems

#### (a) Explain why a transformation was necessary.

\vspace{4cm}

#### (b) Check all model conditions based on the model fit using the transformed data. For any conditions that are not met, suggest a step to take to address the problem.

\newpage

#### (c) Summarize the findings from this analysis about the strength of evidence of an association between body size, gestation length, litter size, and brain weight.

\vspace{8cm}

#### (d) What is the interpretation of the coefficient estimate for the Body variable in the model fit including all observations (Model 1)?

\vspace{4cm}

#### (e) What is the interpretation of the confidence interval for the coefficient of the Body variable in the model fit including all observations (Model 1)?  Include a description of the meaning of the phrase "95% confident".

\vspace{4cm}

#### (f) The variance inflation factor for Body is 3.75.  Rounding up to 4 for convenience, what does this value say about the width of a confidence interval for the coefficient of Body in the linear model?

\vspace{4cm}

#### (g) In the added variable plot for the Body variable, what is on the horizontal and vertical axes of the plot?  How does the slope of the line in that plot relate to the coefficient estimate in the linear model?

\vspace{4cm}

#### (h) What is a high leverage observation?  Why can high leverage observations be problematic?

(You should be able to draw a picture of a simple linear regression setting illustrating why this is a problem.)

\vspace{8cm}

#### (i) What does Cook's distance measure, at an intuitive level?

\newpage

#### (j) Define multicollinearity in a sentence or two.  Why can multicollinearity be problematic?

