---
title: "Chapter 11: Outliers and Influential Observations"
subtitle: "Nov. 11, 2019"
output:
  pdf_document:
    keep_tex: true
    highlight: zenburn
header-includes:
   - \usepackage{soul}
   - \usepackage{booktabs}
documentclass: extarticle
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
library(gridExtra)

anscombe <- read_csv("http://www.evanlray.com/data/base_r/anscombe_quintet.csv")
```

### Recall Anscombe's Data

```{r, echo = FALSE}
fit1 <- lm(y1 ~ x1, data = anscombe)
fit2 <- lm(y2 ~ x2, data = anscombe)
fit3 <- lm(y3 ~ x3, data = anscombe)
fit4 <- lm(y4 ~ x4, data = anscombe)
fit5 <- lm(y5 ~ x5, data = anscombe)

anscombe_long <- data.frame(
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4, anscombe$x5),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4, anscombe$y5),
  residual = c(residuals(fit1), residuals(fit2), residuals(fit3), residuals(fit4), residuals(fit5)),
  data_set = rep(paste("Data Set", 1:5), each = nrow(anscombe))
)

grid.arrange(
  ggplot(data = anscombe_long, mapping = aes(x = x, y = y)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~factor(data_set), nrow = 1) +
    theme_bw(),
  ggplot(data = anscombe_long, mapping = aes(x = x, y = residual)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~factor(data_set), nrow = 1) +
    ylab("residual\n") +
    theme_bw()
)
```

* For today, let's focus on Data Sets 3 and 4.  We will see how to identify the problematic observations from the diagnostics.

* In data set 3, observation 3 is the one with a big Y!

```{r}
anscombe$y3[3]
```

* In data set 4, observation 8 is the one with a big X!

```{r}
anscombe$x4[8]
```

\newpage

## Data Set 3

 * Every statistical software package will give you different plots by default.  Here is my preferred option:

```{r, fig.height = 1.65}
fit3 <- lm(y3 ~ x3, data = anscombe)
anscombe <- anscombe %>%
  mutate(
    obs_index = row_number(),
    h3 = hatvalues(fit3),
    studres3 = rstudent(fit3),
    D3 = cooks.distance(fit3)
  )

# 2p/n; p = 2 since we have beta_0 and beta_1 in our simple linear regression model
2 * 2 / nrow(anscombe)

ggplot(data = anscombe, mapping = aes(x = obs_index, y = h3)) +
  geom_point() +
  geom_hline(yintercept = 2 * 2 / nrow(anscombe)) +
  ggtitle("Leverage - Data Set 3")

ggplot(data = anscombe, mapping = aes(x = obs_index, y = studres3)) +
  geom_point() +
  ggtitle("Studentized Residuals - Data Set 3")

ggplot(data = anscombe, mapping = aes(x = obs_index, y = D3)) +
  geom_point() +
  ggtitle("Cook's Distance - Data Set 3")
```


\newpage

## Data Set 4

```{r, fig.height = 1.55}
fit4 <- lm(y4 ~ x4, data = anscombe)
anscombe <- anscombe %>%
  mutate(
    obs_index = row_number(),
    h4 = hatvalues(fit4),
    studres4 = rstudent(fit4),
    D4 = cooks.distance(fit4)
  )

# 2p/n; p = 2 since we have beta_0 and beta_1 in our simple linear regression model
2 * 2 / nrow(anscombe)

ggplot(data = anscombe, mapping = aes(x = obs_index, y = h4)) +
  geom_point() +
  geom_hline(yintercept = 2 * 2 / nrow(anscombe)) +
  ggtitle("Leverage - Data Set 4")

ggplot(data = anscombe, mapping = aes(x = obs_index, y = studres4)) +
  geom_point() +
  ggtitle("Studentized Residuals - Data Set 4")

ggplot(data = anscombe, mapping = aes(x = obs_index, y = D4)) +
  geom_point() +
  ggtitle("Cook's Distance - Data Set 4")
```

The lower two plots look OK... what's up with that warning?

```{r}
anscombe$h4
anscombe$studres4
anscombe$D4
```

\newpage

## R Code: Default Plots

You can get a set of different diagnostic plots more easily, but I find the plot involving Cook's distance and Leverage less intuitive:

```{r, fig.height = 4, fig.width = 4}
plot(fit3)
```

Note: to get the plots to all show up in the knitted pdf, I had to set figure height and width in the code chunk declaration:

````markdown
`r ''````{r, fig.height = 4, fig.width = 4}
````
