---
title: "R Code: Multiple Comparisons (Sleuth3 Sections 6.3 and 6.4)"
date: "2019-09-25"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
    highlight: zenburn
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
library(ggplot2)
library(readr)
library(dplyr)
library(gmodels)
```

### R code, Handicaps Study

```{r, message = FALSE}
handicaps <- read_csv("http://www.evanlray.com/data/sleuth3/ex0601_handicaps.csv") %>%
  mutate(
    Handicap = factor(Handicap, levels = c("None", "Amputee", "Crutches", "Hearing", "Wheelchair"))
  )

nrow(handicaps)

head(handicaps)

handicaps %>%
  count(Handicap)
```

To keep things shorter, suppose we just want to find 4 confidence intervals:

* $\mu_1 - \mu_2$
* $\mu_1 - \mu_3$
* $\mu_1 - \mu_4$
* $\mu_1 - \mu_5$

### Bonferroni adjustment

#### Step 1: Figure out the individual confidence levels that get you the desired familywise confidence level

 * Can always work through the table
 * General formula is $1 - \alpha/k$ where:
     * $k$ is number of intervals we're making (4 in our example)
     * If we want a 95\% familywise confidence level, $\alpha = 0.05$

```{r}
1 - 0.05/4
```

#### Step 2: Make the confidence intervals

```{r}
anova_fit <- lm(Score ~ Handicap, data = handicaps)
fit.contrast(anova_fit, "Handicap", c(1, -1, 0, 0, 0), conf.int = 0.9875)
fit.contrast(anova_fit, "Handicap", c(1, 0, -1, 0, 0), conf.int = 0.9875)
fit.contrast(anova_fit, "Handicap", c(1, 0, 0, -1, 0), conf.int = 0.9875)
fit.contrast(anova_fit, "Handicap", c(1, 0, 0, 0, -1), conf.int = 0.9875)
```

#### We could also calculate by hand based on the estimates and standard errors from R:

 * The multiplier is the quantile of the t distribution at the point $1 - \alpha/2k$

```{r}
bonferroni_multiplier <- qt(1 - 0.05/(2*4), df = 70 - 5)
bonferroni_multiplier
```

```{r}
c(-0.471 - bonferroni_multiplier * 0.617, -0.471 + bonferroni_multiplier * 0.617)
c(1.021 - bonferroni_multiplier * 0.617, 1.021 + bonferroni_multiplier * 0.617)
c(0.85 - bonferroni_multiplier * 0.617, 0.85 + bonferroni_multiplier * 0.617)
c(-0.443 - bonferroni_multiplier * 0.617, -0.443 + bonferroni_multiplier * 0.617)
```

\newpage

#### Scheffe 95% familywise CIs

 * Use $\text{Multiplier} = \sqrt{(I - 1) F_{(I-1),(n - 1)}(1 - \alpha)}$

First, find the multiplier

```{r}
scheffe_multiplier <- sqrt((5 - 1) * qf(0.95, df1 = 5 - 1, df2 = 70 - 5))
scheffe_multiplier
```

```{r}
c(-0.471 - scheffe_multiplier * 0.617, -0.471 + scheffe_multiplier * 0.617)
c(1.021 - scheffe_multiplier * 0.617, 1.021 + scheffe_multiplier * 0.617)
c(0.85 - scheffe_multiplier * 0.617, 0.85 + scheffe_multiplier * 0.617)
c(-0.443 - scheffe_multiplier * 0.617, -0.443 + scheffe_multiplier * 0.617)
```

#### What is the interpretation of the Scheffe intervals above in context?  As part of your answer, explain what it means that they have a 95% familywise confidence level.

