---
title: "R code: F tests for ANOVA (Sleuth3 Section 5.3)"
subtitle: "We will refine this R code next class"
date: "2019-09-20"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
  - \usepackage{multirow}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r, message = FALSE, echo = FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
options("pillar.sigfig" = 10) # print 10 significant digits in summarize output
```

# Iris Flowers Example

Conduct a claim of the test that the population mean sepal width is the same for all three species.

$H_0: \mu_1 = \mu_2 = \mu_3$

$H_A:$ it is not the case that all three means are equal.

#### Any approach will start off by fitting the full model

```{r, echo = TRUE, fig.height = 1.5, fig.width = 7}
full_fit <- lm(Sepal.Width ~ Species, data = iris)
```

#### Option 1: F statistic and p-value at bottom of model summary

```{r}
summary(full_fit)
```

#### Option 2: Call `anova` on your model fit

```{r}
anova(full_fit)
```

#### Option 3: Fit a reduced model and call `anova`, supplying both the full and reduced models.

For the reduced model, our explanatory variable is "1" -- which means, include only an intercept.

```{r}
reduced_fit <- lm(Sepal.Width ~ 1, data = iris)
anova(reduced_fit, full_fit)
```

This third approach is too much work for the current example, but it is necessary in the examples we will look at next week.
