---
title: "R Examples: Confidence Intervals for the Mean in \"Simple\" Linear Regression"
subtitle: "(Sleuth 3 Sections 7.4.2 and 7.4.3)"
output:
  pdf_document:
    highlight: zenburn
header-includes:
   - \usepackage{soul}
geometry: margin=0.6in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(mosaic)
```

## Goals for today

 * Get an estimate for the mean response at a particular value $x$ of the explanatory variable by plugging in $x$ in the equation
 * Get a *confidence interval* for the mean response at a particular value $x$ using $t$-based methods
 * Make Bonferroni or Scheffe adjustments to get simultaneous confidence intervals for the mean response at multiple values of $x$.

## Example

We have a data set with information about 152 flights by Endeavour Airlines that departed from JFK airport in New York to either Nashville (BNA), Cincinnati (CVG), or Minneapolis-Saint Paul (MSP) in January 2012.

```{r, echo=FALSE, message = FALSE}
flights <- read_csv("flights_data.csv")
```

```{r}
head(flights)
nrow(flights)
```

#### R Code to get model fit

```{r}
model_fit <- lm(air_time ~ distance, data = flights)
summary(model_fit)
```

\newpage

#### Our Estimates for $\beta_0$ and $\beta_1$ are

\vspace{0.5cm}

$\hat{\beta}_0 =$

\vspace{0.5cm}

$\hat{\beta}_1 =$

\vspace{0.5cm}

Note: LaTeX code for $\hat{\beta}_1$ is `\hat{\beta}_1`.

#### Our estimated mean function is

\vspace{0.5cm}

$\hat{\mu}(Y|X) = \hat{\beta}_0 + \hat{\beta}_1 X$ = 

\vspace{0.5cm}

#### The estimated (predicted) mean air time at a flight distance of 589 miles is

\vspace{3cm}

```{r}
14.568 + 0.147 * 589
```

...or...

```{r}
predict_df <- data.frame(
  distance = 589
)
predict_df

predict(model_fit, newdata = predict_df)
```

\newpage

#### Find and interpret a 95% confidence interval for the mean air time for flights traveling a distance of 589 miles.

```{r}
predict(model_fit, newdata = predict_df, interval = "confidence", se.fit = TRUE)
```

\vspace{8cm}

This is just a $t$-based interval based on the estimate and its standard error (although the calculation of standard error is complicated...)

```{r}
qt(0.975, df = 152 - 2)

101.150 - 1.976 * 1.526
101.150 + 1.976 * 1.526
```

\newpage

#### Find and interpret Bonferroni adjusted confidence intervals for the mean air time at flight distances of 589 miles, 765 miles, and 1029 miles, with a familywise confidence level of 95%.

Approach 1 (easier): adjust confidence level we ask `predict` for.

 * 3 CI's at a familywise confidence level of 95%
 * Overall, miss for 5% of samples, $\alpha = 0.05$
 * Each individual CI has $\alpha = 0.05/3 = 0.0167$
 * Each individual CI has confidence level $(1 - 0.0167) \times 100\% = 98.3\%$

```{r}
predict_df <- data.frame(
  distance = c(589, 765, 1029)
)
predict_df

predict(model_fit,
  newdata = predict_df,
  interval = "confidence",
  se.fit = TRUE,
  level = 0.983
)
```

\vspace{2cm}

**Approach 2 (you don't ever need to do this): Manual calculation based on standard errors**

```{r}
(1 - 0.05/(2*3))
qt(0.9917, df = 152 - 2)
```

```{r}
# CI for X0 = 589 -- the other two are calculated similarly
101.150 - 2.423 * 1.526
101.150 + 2.423 * 1.526
```

\newpage

#### Find and plot Scheffe adjusted CIs for the means at a grid of 100 values of x between 589 and 1029

```{r, fig.height = 8.5, message = FALSE}
library(lava) # contains the scheffe function
predict_df <- data.frame(
  distance = seq(from = 589, to = 1029, length = 100)
)
head(predict_df, 3)
scheffe_cis <- scheffe(model_fit, predict_df)
head(scheffe_cis, 3)

predict_df <- predict_df %>% mutate(
  scheffe_lwr = scheffe_cis[, 2],
  scheffe_upr = scheffe_cis[, 3]
)
head(predict_df, 3)

ggplot(data = flights, mapping = aes(x = distance, y = air_time)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_lwr)) +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_upr)) +
  theme_bw()
```

 * The upper black line shows the upper bounds for Scheffe-adjusted confidence intervals for the mean at 100 different distances, connected together (similar for lower)
 * The grey shaded region shows unadjusted confidence intervals
 * Interpretation:
    * We are 95% confident that at every distance, the population mean air time at that distance is between the lines.
    * Each sample would give us a different set of lines
    * For 95% of samples, the population mean air time would be between the lines **at every distance** (if all conditions are satisfied).

\newpage

#### What if we had used Bonferroni intervals at each of the 100 distances?

 * To get 95% familywise confidence level, each CI needs to have confidence level 99.95%

```{r, fig.height = 5.25}
bonferroni_cis <- predict(model_fit,
  newdata = predict_df,
  interval = "confidence",
  level = 0.9995
)

predict_df <- predict_df %>% mutate(
  bonferroni_lwr = bonferroni_cis[, 2],
  bonferroni_upr = bonferroni_cis[, 3]
)

ggplot(data = flights, mapping = aes(x = distance, y = air_time)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_lwr)) +
  geom_line(data = predict_df, mapping = aes(x = distance, y = scheffe_upr)) +
  geom_line(data = predict_df, mapping = aes(x = distance, y = bonferroni_lwr), color = "orange") +
  geom_line(data = predict_df, mapping = aes(x = distance, y = bonferroni_upr), color = "orange") +
  theme_bw()
```

 * Bonferroni:
    * The more confidence intervals you're making, the wider Bonferroni-adjusted intervals are
 * Scheffe:
    * Intervals are always the same width, no matter how many you make
