---
title: "$t$ Tests Involving a Linear Combination of Multiple Means"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

#### The ANOVA Model

 * We have $I$ groups ($I = 3$ for iris example)
 * Sample size of $n_i$ for group $i$, total sample size $n = n_1 + n_2 + \cdots + n_I$
 * Observations in group $i$ follow a Normal($\mu_i$, \sigma^2) distribution
    * (Potentially) different mean for each group
    * Same variance across all groups

Theoretical model:

```{r, echo = FALSE, fig.height = 2}
x_grid <- seq(from = 0, to = 20, length = 101)
group_sd <- 2

plot_df <- data.frame(
  x = rep(x_grid, 3),
  y = c(dnorm(x_grid, mean = 8, sd = group_sd),
        dnorm(x_grid, mean = 10, sd = group_sd),
        dnorm(x_grid, mean = 14, sd = group_sd)),
  group = rep(c("Group 1: mean = 8, sd = 2", "Group 2: mean = 10, sd = 2", "Group 3: mean = 14, sd = 2"),
    each = length(x_grid))
)

ggplot(data = plot_df, mapping = aes(x = x, y = y, color = group)) +
  geom_line()
  
```

Compare to the plot for our iris data:

```{r, fig.height = 2}
ggplot(data = iris, mapping = aes(x = Sepal.Width, color = Species)) +
  geom_density()
```

#### Hypothesis

 * General case:
 
$H_0: C_1 \mu_1 + C_2 \mu_2 + \cdots + C_I \mu_I = 0$

$H_A: C_1 \mu_1 + C_2 \mu_2 + \cdots + C_I \mu_I \neq 0$

 * Example: Test of whether $\mu_1 = \mu_2$, or equivalently $\mu_1 - \mu_2 = 0$

$H_0: 1 \times \mu_1 + (-1) \times \mu_2 + 0 \times \mu_3 + \cdots + 0 \times \mu_I = 0$

$H_A: 1 \times \mu_1 + (-1) \times \mu_2 + 0 \times \mu_3 + \cdots + 0 \times \mu_I \neq 0$

$C_1 = 1$, $C_2 = -1$, $C_3 = 0$, $\ldots$, $C_I = 0$



 * The variance the average squared difference from the mean.
 
 * Imagine a population of $N$ people, each has a height $Y_i$
 
 * Mean height is $\mu = \frac{1}{N} \sum_{i = 1}^N Y_i$
 
 * Variance of heights is $\sigma^2 = \frac{1}{N} \sum_{i = 1}^N (Y_i - \mu)^2$

```{r, echo = FALSE, fig.height = 3, message = FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(mosaic)

ggplot(data = data.frame(Y = c(-10, 10)), mapping = aes(x = Y)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 4), color = "orange") +
  theme_bw()
```

 * Both distributions have mean 0
 * The distribution shown in orange has larger variance than the distribution shown in black
    * A typical value is farther from 0
    * A typical value has larger squared distance from 0
    * Across all values, the average squared difference from 0 is larger

#### Sample Variance

 * The *sample variance* $s^2$ is used as an estimate of the population variance $\sigma^2$.

 * Suppose we tried to use

$\tilde{s}^2 = \frac{1}{n}\sum_{i = 1}^n(Y_i - \bar{Y})^2$

 * Doesn't work because:
    * Values in sample tend to be a little closer to $\bar{Y}$ than to $\mu$
    * Squared differences from $\bar{Y}$ are a little smaller than squared differences from $\mu$
    * $\tilde{s}^2$ tends to be less than $\sigma^2$

 * Dividing by $n - 1$ instead of $n$ is just the right adjustment:

$s^2 = \frac{1}{n - 1}\sum_{i = 1}^n(Y_i - \bar{Y})^2$

\newpage

#### Demonstration by simulation

 * "Population": Normal$(0, 1)$  Variance is $\sigma^2 = 1$

 * Sample of size $n = 4$

```{r, echo = FALSE, fig.height = 1.7}
set.seed(847)

sample_df <- data.frame(
  x = rnorm(4),
  y = 0
)

ggplot(data = data.frame(x = c(-4, 4)), mapping = aes(x = x)) + 
  stat_function(fun = dnorm) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = mean(sample_df$x), color = "orange", linetype = 2, size = 1) +
  geom_point(data = sample_df, mapping = aes(y = y), color = "orange", size = 3) +
  ylab("") +
  ggtitle("In black: Population distribution, population mean\nIn orange: Sample observations, sample mean") +
  theme_bw()

sample_df <- sample_df %>% select(x)
```

Here's our data frame with sample data:

```{r}
sample_df
```

Sample mean is:

```{r}
sample_df %>%
  summarize(
    mean = mean(x)
  )
```

Average squared difference from population mean of 0:

```{r}
sample_df %>%
  summarize(
    mean_squared_difference_from_0 = mean((x - 0)^2)
  )
```

Average squared difference from sample mean of 0.376:

```{r}
sample_df %>%
  summarize(
    mean_squared_difference_from_0 = mean((x - 0.376)^2)
  )
```

Corrected by dividing by $n - 1$ instead of $n$:

```{r}
sample_df %>%
  summarize(
    almost_mean_squared_difference_from_0 = sum((x - 0.376)^2) / (4 - 1)
  )
```

 * For this particular sample, all three are below the population variance of 1.  What about across 10000 samples?

```{r, echo = FALSE, cache = TRUE}
sample_df <- data.frame(
  x = rnorm(4),
  y = 0
)

results <- do(10000) * {
  sample_df <- data.frame(
    x = rnorm(4)
  )
  
  data.frame(
    `Denominator n` = mean((sample_df$x - mean(sample_df$x))^2),
    `Denominator n minus 1` = sum((sample_df$x - mean(sample_df$x))^2) / (3)
  )
}
```

```{r, echo = FALSE}
results <- results %>%
  select(1:2) %>%
  gather(
    "denominator", value = "variance_estimate"
  ) %>%
  mutate(
    denominator = ifelse(denominator == "Denominator.n", "Denominator n", "Denominator n-1")
  )
```

```{r, echo = FALSE}
ggplot(data = results, mapping = aes(x = variance_estimate)) +
  geom_histogram(boundary = 0, fill = "orange", alpha = 0.5, bins = 30) +
  geom_vline(xintercept = 1) +
  geom_vline(results %>% group_by(denominator) %>% summarize(mean_est = mean(variance_estimate)),
    mapping = aes(xintercept = mean_est), color = "orange", linetype = 2, size = 1) +
  facet_wrap(~ denominator, ncol = 1) +
  ggtitle("Black line: population variance\nOrange line: average estimate of population variance across 10000 samples") +
  theme_bw()
```

#### Connection to $t$ tests

 * Our $t$ statistic for a test about the value of a mean $\mu$ is

$t = \frac{\bar{Y} - \mu}{SE(\bar{Y})} = \frac{\bar{Y} - \mu}{s / \sqrt{n}}$

 * $s$ shows up in this calculation, with associated degrees of freedom $n - 1$.


