---
title: "ANOVA: First Examples, Model Statement, t tests"
subtitle: "Sleuth3 Sections 6.2 and 5.2"
output:
  pdf_document:
    fig_height: 2.8
    fig_width: 6
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes:
  - \usepackage{booktabs}
geometry: margin=0.6in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Examples, First Look at Data

## Example 1: Sepal Width of Iris Flowers

#### Study Overview

We have measurements of the characteristics of 150 iris flowers, 50 each of three different species:

 * Iris setosa is found in the arctic, including Alaska and Maine in the United States, Canada, Russia, northern China, Korea and other northern countries.
 * Iris versicolor is found in the eastern United States and eastern Canada.
 * Iris virginica is found in the eastern United States

It's not clear how the flowers were selected for the sample; probably not as a representative sample though.  The original purpose of this study was to develop methods for identifying a flower's species based on physical measurements of its characteristics.  One of the characteristics that was measured for each flower in our sample was the width of the flower's sepal; the sepal is the part of the plant that sits just below the petals and supports them.

We will investigate differences in the widths of the flowers' sepals for the different species.

#### Look at the Data:

```{r, message = FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
options("pillar.sigfig" = 10) # print 10 significant digits in summarize output

head(iris)
dim(iris)

iris %>%
  count(Species)
```

\newpage

General template of code to make plots with `ggplot2`:

```{r, eval = FALSE}
ggplot(data = <name of data frame>,
  mapping = aes(x = <variable for x axis>,
    y = <variable for y axis>,
    color = <variable for color lines>,
    fill = <variable for color area>,
    )) +
  geom_<geometry type>() +
  <optional other things like facetting, axis labels, ...>
```

One pair of figures to sum up ANOVA:

 * **Key idea** Less variability within each group than across all flowers

```{r, fig.height = 1.5}
ggplot(data = iris, mapping = aes(x = Sepal.Width)) +
  geom_histogram(binwidth = 0.2)
```

```{r, fig.height = 4.5}
ggplot(data = iris, mapping = aes(x = Sepal.Width)) +
  geom_histogram(binwidth = 0.2) +
  facet_wrap( ~ Species, ncol = 1)
```

\newpage

Box plots:

```{r, fig.height=2}
plot_combined <- ggplot(data = iris, mapping = aes(y = Sepal.Width)) +
  geom_boxplot() +
  xlab("All groups combined")

plot_bygroup <- ggplot(data = iris, mapping = aes(y = Sepal.Width, x = Species)) +
  geom_boxplot()

grid.arrange(plot_combined, plot_bygroup, ncol = 2)
```


Same idea, but with density plots instead of histograms:

```{r, fig.height = 2, fig.width = 4.8}
ggplot(data = iris, mapping = aes(x = Sepal.Width)) +
  geom_density()
```

```{r, fig.height = 2}
ggplot(data = iris, mapping = aes(x = Sepal.Width, color = Species)) +
  geom_density()
```

\newpage

```{r}
# Calculate overall sample mean and standard deviation
iris %>%
  summarize(
    mean = mean(Sepal.Width),
    sd = sd(Sepal.Width)
  )

  summarize(
    iris,
    mean = mean(Sepal.Width),
    sd = sd(Sepal.Width)
  )



# Calculate sample means and standard deviations separately for each species
iris %>%
  group_by(Species) %>%
  summarize(
    mean = mean(Sepal.Width),
    sd = sd(Sepal.Width)
  )
```

#### Parameters:

$\mu_{1}$ = Average sepal width among all setosa flowers (in the region where the flowers in the sample were found?)

$\mu_{2}$ = Average sepal width among all versicolor flowers (in the region where the flowers in the sample were found?)

$\mu_{3}$ = Average sepal width among all virginica flowers (in the region where the flowers in the sample were found?)

#### Some questions we might ask:

 1. How strong is the evidence against the claim that the means are the same for setosa flowers and versicolor flowers?
    * $H_0:$
 2. How strong is the evidence against the claim that the means are the same for setosa flowers and virginica flowers?
    * $H_0:$
 3. How strong is the evidence against the claim that the means are the same for versicolor flowers and virginica flowers?
    * $H_0:$
 4. How strong is the evidence against the claim that the mean for setosa flowers, found in the arctic, is the same as the mean for non-arctic flowers?
    * $H_0:$
 5. Overall, how strong is the evidence against the claim that there is no difference in means for the three species?
    * $H_0:$

\newpage


# ANOVA Model, Hypothesis Tests, and Confidence Intervals

## The ANOVA Model

 * We have $I$ groups ($I = 3$ for iris example)
 * Sample size of $n_i$ for group $i$, total sample size $n = n_1 + n_2 + \cdots + n_I$
 * Observations in group $i$ follow a Normal($\mu_i$, $\sigma^2$) distribution
    * (Potentially) different mean for each group
    * Same variance across all groups
 * All observations are independent of each other: knowing that one is above its group mean doesn't tell you whether or not another is above its group mean.

#### Theoretical model:

```{r, echo = FALSE, fig.height = 1.5}
x_grid <- seq(from = 0, to = 20, length = 101)
group_sd <- 2

plot_df <- data.frame(
  x = rep(x_grid, 3),
  y = c(dnorm(x_grid, mean = 8, sd = group_sd),
        dnorm(x_grid, mean = 10, sd = group_sd),
        dnorm(x_grid, mean = 14, sd = group_sd)),
  group = rep(c("Group 1: mean = 8, sd = 2", "Group 2: mean = 10, sd = 2", "Group 3: mean = 14, sd = 2"),
    each = length(x_grid))
)

ggplot(data = plot_df, mapping = aes(x = x, y = y, color = group)) +
  geom_line()
```

#### Compare to the plot for our iris data:

```{r, fig.height = 1.5}
ggplot(data = iris, mapping = aes(x = Sepal.Width, color = Species)) +
  geom_density()
```

## Hypotheses

#### General case:

$H_0: C_1 \mu_1 + C_2 \mu_2 + \cdots + C_I \mu_I = 0$

$H_A: C_1 \mu_1 + C_2 \mu_2 + \cdots + C_I \mu_I \neq 0$

 * Notation: Book defines $\gamma$ ("gamma") to be the linear combination we are testing:

$\gamma = C_1 \mu_1 + C_2 \mu_2 + \cdots + C_I \mu_I$

#### Iris example:

State the hypotheses and the values of the constants $C_{1}$, $C_2$, and $C_3$ for a test of the claim that the mean for setosa flowers is equal to the mean for versicolor flowers.

\vspace{5cm}

\newpage

## Doing the test and finding a 95% confidence interval

Here's some R code to do this test

```{r}
library(gmodels)

model_fit <- lm(Sepal.Width ~ Species, data = iris)
fit.contrast(model_fit, "Species", c(1, -1, 0), conf = 0.95)
```

#### What is the result of the hypothesis test?  State a conclusion in terms of strength of evidence against the null hypothesis.

\vspace{5cm}

#### What is the interpretation of the confidence interval?  In your answer, include a statement of what the phrase "95% Confident" means.

\vspace{5cm}

#### Suppose you want to conduct a hypothesis test of whether the mean for the setosa flower, found in the arctic, is the same as the mean across both non-arctic flowers.  What constants $C_1$, $C_2$, and $C_3$ would you use?

The null hypothesis is $H_0: \mu_{1} = \frac{1}{2}(\mu_{2} + \mu_{3})$

\newpage

## Example 2: Women underrepresented on juries?

Quote from our book:

"In 1968, Dr. Benjamin Spock was tried in Boston on charges of conspiring to violate the Selective Service Act by encouraging young men to resist being drafted into military service for Vietnam. The defence in the case challenged the method of jury selection claiming that women were underrepresented. Boston juries are selected in three stages. First 300 names are selected at random from the City Directory, then a venire of 30 or more jurors is selected from the initial list of 300 and finally, an actual jury is selected from the venire in a nonrandom process allowing each side to exclude certain jurors. There was one woman on the venire and no women on the final list. The defence argued that the judge in the trial had a history of venires in which women were systematically underrepresented and compared the judge's recent venires with the venires of six other Boston area district judges."

```{r, message = FALSE}
library(readr)
library(ggplot2)
library(dplyr)

juries <- read_csv("http://www.evanlray.com/data/sleuth3/ex0502_women_jurors.csv")
```

```{r}
dim(juries)
head(juries)

juries %>% count(Judge)
```

Each observational unit is a venire (jury pool) assembled by one of 7 judges in Boston at this time.  We have information about a total of 46 venires across the 7 judges.  For each venire, we have recorded:

 * The percent of potential jurors in the venire who were women
 * Which judge assembled that venire

\newpage

#### Initial Plots

The GitHub repository URL for this lab is: `https://github.com/mhc-stat242-s2019/Lab2.git`

In R, recreate at least one of the plots below.  Also calculate the group means and standard deviations.

```{r, fig.height = 2, echo = FALSE}
ggplot(data = juries, mapping = aes(x = Judge, y = Percent)) +
  geom_boxplot() +
  theme_bw()
```

```{r, fig.height=2.5, echo = FALSE}
ggplot(data = juries, mapping = aes(x = Percent)) +
  geom_histogram(binwidth = 2.5) +
  facet_wrap( ~ Judge, ncol = 4) +
  theme_bw()
```

\newpage

## Example 2

#### (a) Conduct a hypothesis test of the claim that the mean percent of potential jurors who are women in venires assembled by Spock's judge is the same as the mean percent of potential jurors who are women in venires assembled by judge A.  Also find and report a 95% confidence interval for the difference in means for those two judges.  State your null and alternative hypotheses in terms of equations and written sentences.  What are the constants $C_1, \ldots, C_I$ to use for this procedure?

\vspace{8cm}

#### (b) Conduct a hypothesis test of the claim that the mean percent of potential jurors who are women in venires assembled by Spock's judge is the same as the mean percent of potential jurors who are women across all 6 other judges.  Also find and report a 95% confidence interval for the difference in means between Spock's judge and the average across all 6 other judges.  State your null and alternative hypotheses in terms of equations and written sentences.  What are the constants $C_1, \ldots, C_I$ to use for this procedure?

